---
title: "May 2023"
author: 
  - name: "Tomasz Woźniak"
    url: https://github.com/donotdespair
    affiliations:
      - University of Melbourne
    orcid: 0000-0003-2212-2378
execute:
  freeze: auto
---

```{r interest data}
#| echo: false
#| message: false
#| warning: false

# download daily interest rates
icr_dwnld   = readrba::read_rba(series_id = "FIRMMCRTD")   # Cash Rate Target
icr_tmp     = xts::xts(icr_dwnld$value, icr_dwnld$date)
dates_tmp   = xts::xts(as.Date(icr_dwnld$date), icr_dwnld$date)

by1m_dwnld = readrba::read_rba(series_id = "FIRMMBAB30D")
by1m_tmp   = xts::xts(by1m_dwnld$value, by1m_dwnld$date)

by3m_dwnld = readrba::read_rba(series_id = "FIRMMBAB90D")
by3m_tmp   = xts::xts(by3m_dwnld$value, by3m_dwnld$date)

by6m_dwnld = readrba::read_rba(series_id = "FIRMMBAB180D")
by6m_tmp   = xts::xts(by6m_dwnld$value, by6m_dwnld$date)

by2y_dwnld = readrba::read_rba(series_id = "FCMYGBAG2D")
by2y_tmp   = xts::xts(by2y_dwnld$value, by2y_dwnld$date)

by3y_dwnld = readrba::read_rba(series_id = "FCMYGBAG3D")
by3y_tmp   = xts::xts(by3y_dwnld$value, by3y_dwnld$date)

by5y_dwnld = readrba::read_rba(series_id = "FCMYGBAG5D")
by5y_tmp   = xts::xts(by5y_dwnld$value, by5y_dwnld$date)

by10y_dwnld = readrba::read_rba(series_id = "FCMYGBAG10D")
by10y_tmp   = xts::xts(by10y_dwnld$value, by10y_dwnld$date)

# daily systems

forecast_day              = "/2023-05-31"

variables_all             = na.omit(merge(icr_tmp, by1m_tmp, by3m_tmp, by6m_tmp, by2y_tmp, by3y_tmp, by5y_tmp, by10y_tmp))
colnames(variables_all)   = c("cash rate", "1m", "3m", "6m", "2y", "3y", "5y", "10y")
variables_all             = variables_all[forecast_day]

variables_long            = na.omit(merge(icr_tmp, by2y_tmp, by3y_tmp, by5y_tmp, by10y_tmp))
colnames(variables_long)  = c("cash rate", "2y", "3y", "5y", "10y")
variables_long            = variables_long[forecast_day]

variables_short           = na.omit(merge(icr_tmp, by1m_tmp, by3m_tmp, by6m_tmp))
colnames(variables_short) = c("cash rate", "1m", "3m", "6m")
variables_short           = variables_short[forecast_day]


# weekly and monthly systems
vwa     = xts::to.weekly(variables_all, OHLC = FALSE)
vwl     = xts::to.weekly(variables_long, OHLC = FALSE)
vws     = xts::to.weekly(variables_short, OHLC = FALSE)

vma     = xts::to.monthly(variables_all, OHLC = FALSE)
vml     = xts::to.monthly(variables_long, OHLC = FALSE)
vms     = xts::to.monthly(variables_short, OHLC = FALSE)

# create a dummy for the interest raise regime
T       = nrow(vms)
dm      = xts::xts(as.matrix(rep(0, T)), zoo::index(vms))
dm["2022-05/"] = 1
colnames(dm) = "dum"
dmf = as.matrix(rep(1,12))
colnames(dmf) = "dum"

Tw      = nrow(vws)
dw      = xts::xts(as.matrix(rep(0, Tw)), zoo::index(vws))
dw["2022-05/"] = 1
colnames(dw) = "dum"
dwf = as.matrix(rep(1,55))
colnames(dwf) = "dum"
```

```{r cointegrating rank}
#| echo: false
#| eval: false
#| message: false
#| warning: false
#| results: hide

library(vars)
# Johansen's cointegrating rank test
vecm_vma   = ca.jo(vma, type = "trace", ecdet = "const", K = 5, spec = "transitory")
summary(vecm_vma) # r = 5, N = 8

vecm_vms   = ca.jo(vms, type = "trace", ecdet = "const", K = 5, spec = "transitory")
summary(vecm_vms) # r = 3, N = 4

vecm_vml   = ca.jo(vml, type = "trace", ecdet = "const", K = 5, spec = "transitory")
summary(vecm_vml) # r = 4, N = 5


vecm_vwa   = ca.jo(vwa, type = "trace", ecdet = "const", K = 5, spec = "transitory")
summary(vecm_vwa) # r = 7, N = 8

vecm_vws   = ca.jo(vws, type = "trace", ecdet = "const", K = 5, spec = "transitory")
summary(vecm_vws) # r = 3, N = 4

vecm_vwl   = ca.jo(vwl, type = "trace", ecdet = "const", K = 5, spec = "transitory")
summary(vecm_vwl) # r = 4, N = 5
```

```{r forecasting}
#| echo: false
#| message: false
#| warning: false

library(vars)
# forecast with monthly data
f         = 1
# forecasts = array(NA, c(12, 3, 18))
# loglik    = rep(NA, 18)
forecasts = array(NA, c(12, 3, 16))
loglik    = rep(NA, 16)
vm        = list(vma, vms, vml)
rr        = c(4, 3, 3)

# for (v in 1:3) {
for (v in 2) {
  for (p in c(3, 5, 7, 9, 11, 13, 15, 17)) {
    vecm            = ca.jo(vm[[v]], type = "trace", ecdet = "const", K = p, spec = "transitory", dumvar = dm)
    var_cr          = vec2var(vecm, r = rr[v])
    var_pr          = predict(var_cr, n.ahead = 12, ci = .68, dumvar = dmf)
    forecasts[,,f]  = var_pr$fcst$cash.rate[,1:3]
    loglik[f]       = sum(dnorm(var_cr$resid[,1], log = TRUE))
    f = f + 1
    
    vecm            = ca.jo(vm[[v]], type = "trace", ecdet = "const", K = p, spec = "transitory")
    var_cr          = vec2var(vecm, r = rr[v])
    var_pr          = predict(var_cr, n.ahead = 12, ci = .68)
    forecasts[,,f]  = var_pr$fcst$cash.rate[,1:3]
    loglik[f]       = sum(dnorm(var_cr$resid[,1], log = TRUE))
    f = f + 1
  }
}

ym1 = zoo::as.yearmon("2023-06") # the first forecasted period
ym2 = zoo::as.yearmon("2024-05") # the last forecasted period
s   = seq(ym1, ym2, 1/12) # create yearmon sequence

# weights are proportional to marginal likelihood for the cash rate
ll        = exp(loglik - max(loglik))
weights   = ll/sum(ll)

forecasts_w = forecasts
# for (i in 1:18) {
for (i in 1:16) {
  forecasts_w[,,i]      = weights[i] * forecasts[,,i]
}

# pooled_forecasts_m = apply(forecasts, 1:2, mean)
# colnames(pooled_forecasts_m) = c("forecast", "lower", "upper")
# pooled_forecasts_m = xts::xts(pooled_forecasts_m, s)

pooled_forecasts_m = apply(forecasts_w, 1:2, sum)
colnames(pooled_forecasts_m) = c("forecast", "lower", "upper")
pooled_forecasts_m = xts::xts(pooled_forecasts_m, s)

ym13 = zoo::as.yearmon("2023-05") # forecast origin
s3   = seq(ym13, ym2, 1/12) # create yearmon sequence

ym12 = zoo::as.yearmon("2010-1") # first data point for the plot
s2   = seq(ym12, ym2, 1/12) # create yearmon sequence

datainforecast  = as.vector(vm[[2]][(dim(vm[[2]])[1] - (length(s2) - 12 - 1)):dim(vm[[2]])[1], 1])
last_point      = datainforecast[length(datainforecast)]

cols = c("darkorchid4","mediumorchid1","mediumorchid2","mediumorchid3","hotpink1","hotpink2","hotpink3","hotpink4")



# forecast with weekly data
f           = 1
# forecastsw  = array(NA, c(55, 3, 18))
# loglikw     = rep(NA, 18)
forecastsw  = array(NA, c(55, 3, 16))
loglikw     = rep(NA, 16)
vw          = list(vwa, vws, vwl)
rrw         = c(7, 3, 4)

# for (v in 1:3) {
for (v in 2) {
  for (p in c(3, 5, 7, 9, 11, 13, 15, 17)) {
    vecm            = ca.jo(vw[[v]], type = "trace", ecdet = "const", K = p, spec = "transitory", dumvar = dw)
    var_cr          = vec2var(vecm, r = rrw[v])
    var_pr          = predict(var_cr, n.ahead = 55, ci = .68, dumvar = dwf)
    forecastsw[,,f] = var_pr$fcst$cash.rate[,1:3]
    loglikw[f]      = sum(dnorm(var_cr$resid[,1], log = TRUE))
    f = f + 1
    
    vecm            = ca.jo(vw[[v]], type = "trace", ecdet = "const", K = p, spec = "transitory")
    var_cr          = vec2var(vecm, r = rrw[v])
    var_pr          = predict(var_cr, n.ahead = 55, ci = .68)
    forecastsw[,,f] = var_pr$fcst$cash.rate[,1:3]
    loglikw[f]      = sum(dnorm(var_cr$resid[,1], log = TRUE))
    f = f + 1
  }
}

sw    = as.Date(rep(NA, 55))
sw[1] = as.Date("2023-06-07")
for (i in 2:55) {
  sw[i] = sw[i - 1] + 7
}

# weights are proportional to marginal likelihood for the cash rate
llw        = exp(loglik - max(loglik))
weightsw   = llw/sum(llw)

forecastsw_w = forecastsw
# for (i in 1:18) {
for (i in 1:16) {
  forecastsw_w[,,i]      = weightsw[i] * forecastsw[,,i]
}

pooled_forecasts_ww = apply(forecastsw_w, 1:2, sum)
colnames(pooled_forecasts_ww) = c("forecast", "lower", "upper")
pooled_forecasts_ww = xts::xts(pooled_forecasts_ww, sw)


# pooled_forecasts_w = apply(forecastsw, 1:2, mean)
# colnames(pooled_forecasts_w) = c("forecast", "lower", "upper")
# pooled_forecasts_w = xts::xts(pooled_forecasts_w, sw)
# pooled_forecasts_wm = xts::to.monthly(pooled_forecasts_w, OHLC = FALSE)

pooled_forecasts_wm = xts::to.monthly(pooled_forecasts_ww, OHLC = FALSE)

```

```{r univariate forecasting}
#| echo: false
#| message: false
#| warning: false

crm = vms[,1]
crw = vws[,1]

# monthly arma-garch forecasting
p_max     = 3
q_max     = 3

garchfm   = array(NA, c(12, 3, 4, (p_max + 1) * (q_max + 1)))
loglik    = matrix(NA, 4, (p_max + 1) * (q_max + 1))

for (i in 0:p_max){
  for (j in 0:q_max){
    # arma-garch forecasting
    garch_ms   = rugarch::ugarchspec(
      variance.model = list(model = "sGARCH"), 
      mean.model = list(armaOrder = c(i, j)), 
      distribution.model = "norm"
    )
    garch_mf   = rugarch::ugarchfit(garch_ms, crm, solver = "hybrid")
    garch_mfor = rugarch::ugarchboot(garch_mf, n.ahead = 12,
                                     method = "Partial", n.bootpred = 5000)
    garchfm[,,1, (p_max + 1)*i + j + 1]  = cbind(garch_mfor@forc@forecast$seriesFor,
          t(apply(garch_mfor@fseries, 2, HDInterval::hdi , credMass = .68)))
    loglik[1, (p_max + 1)*i + j + 1]     = -sum(garch_mf@fit$log.likelihoods)
    
    # arma-garch-in-mean forecasting
    garch_ms   = rugarch::ugarchspec(
      variance.model = list(model = "sGARCH"), 
      mean.model = list(armaOrder = c(i, j), archm = TRUE), 
      distribution.model = "norm"
    )
    garch_mf   = rugarch::ugarchfit(garch_ms, crm, solver = "hybrid")
    garch_mfor = rugarch::ugarchboot(garch_mf, n.ahead = 12,
                                     method = "Partial", n.bootpred = 5000)
    garchfm[,,2, (p_max + 1)*i + j + 1]  = cbind(garch_mfor@forc@forecast$seriesFor,
          t(apply(garch_mfor@fseries, 2, HDInterval::hdi , credMass = .68)))
    loglik[2, (p_max + 1)*i + j + 1]     = -sum(garch_mf@fit$log.likelihoods)
    
    # arma-gjr-garch forecasting
    garch_ms   = rugarch::ugarchspec(
      variance.model = list(model = "gjrGARCH"), 
      mean.model = list(armaOrder = c(i, j)), 
      distribution.model = "norm"
    )
    garch_mf   = rugarch::ugarchfit(garch_ms, crm, solver = "hybrid")
    garch_mfor = rugarch::ugarchboot(garch_mf, n.ahead = 12,
                                     method = "Partial", n.bootpred = 5000)
    garchfm[,,3, (p_max + 1)*i + j + 1]  = cbind(garch_mfor@forc@forecast$seriesFor,
          t(apply(garch_mfor@fseries, 2, HDInterval::hdi , credMass = .68)))
    loglik[3, (p_max + 1)*i + j + 1]     = -sum(garch_mf@fit$log.likelihoods)
    
    # arma-gjr-garch-in-mean forecasting
    garch_ms   = rugarch::ugarchspec(
      variance.model = list(model = "gjrGARCH"), 
      mean.model = list(armaOrder = c(i, j), archm = TRUE), 
      distribution.model = "norm"
    )
    garch_mf   = rugarch::ugarchfit(garch_ms, crm, solver = "hybrid")
    garch_mfor = rugarch::ugarchboot(garch_mf, n.ahead = 12,
                                     method = "Partial", n.bootpred = 5000)
    garchfm[,,4, (p_max + 1)*i + j + 1]  = cbind(garch_mfor@forc@forecast$seriesFor,
          t(apply(garch_mfor@fseries, 2, HDInterval::hdi , credMass = .68)))
    loglik[4, (p_max + 1)*i + j + 1]     = -sum(garch_mf@fit$log.likelihoods)
  }
}
loglik[is.na(loglik)] = -1000000

llm        = exp(loglik - max(loglik))
weightsm   = llm/sum(llm)

forecastsm = garchfm
forecastsm[is.na(forecastsm)] = 0
for (i in 1:dim(garchfm)[3]) {
  for (j in 1:dim(garchfm)[4]){
    forecastsm[,,i,j]      = weightsm[i,j] * forecastsm[,,i,j]
  }
}

pooled_forecasts_garchm = apply(forecastsm, 1:2, sum)
colnames(pooled_forecasts_garchm) = c("forecast", "lower", "upper")
pooled_forecasts_garchm = xts::xts(pooled_forecasts_garchm, s)

```

```{r pool forecasts}
#| echo: false
#| message: false
#| warning: false

# pool forecasts
pooled_forecasts = (1/3) * (pooled_forecasts_m + pooled_forecasts_wm + pooled_forecasts_garchm)
zoo::write.zoo(pooled_forecasts, sep = ",", file = "forecasts/2023-05.csv")

```



> The end-of-May forecasting for the RBA cash rate survey by [finder.com.au](https://www.finder.com.au/rba-cash-rate) follows the anouncement of inflation for May 2023 at the level of 6.8%. It fell, but remains high. The new data leads to forecasts indicating a decisive increase in the cash rate in June and subsequent increases further on.

## Cash rate forecasts

The figure below presents the monthly cash rate series starting from January 2010, with the forecasts reported from June 2023 to May 2024 as the forecast mean and the 68% forecasting intervals.

```{r forecast plot}
#| echo: false
#| message: false
#| warning: false

ci1_tmp = col2rgb(cols[2])
ci2_tmp = col2rgb(cols[4])
ci1     = rgb(ci1_tmp[1], ci1_tmp[2], ci1_tmp[3], 100, maxColorValue = 255)
ci2     = rgb(ci2_tmp[1], ci2_tmp[2], ci2_tmp[3], 100, maxColorValue = 255)

plot(x = s2, y = c(datainforecast, pooled_forecasts[,1]), main = "Cash rate forecast",
     type = "l", ylab = "[%]", xlab = "time",
     ylim = range(pooled_forecasts, datainforecast), bty = "n",
     lwd = 1, col = cols[1]
)
polygon(x = c(s3, s3[13:1]), 
        y = c(last_point,as.vector(pooled_forecasts[,2]), as.vector(pooled_forecasts[,3])[12:1], last_point),
        col = ci1, border = ci1)
lines(x = s2, y = c(datainforecast, pooled_forecasts[,1]), lwd = 2, col = cols[1])
abline(v = ym13, col = cols[6], lty = 3)


```

The table below makes the numerical values presented in the figure more accessible.

```{r forecast table}
#| echo: false
#| message: false
#| warning: false
options(knitr.kable.NA = '') 
# pooled_mq     = merge(pooled_forecasts,pooled_forecasts_q)
# colnames(pooled_mq) = c("monthly", "lower", "upper", "quarterly", "lower", "upper")
# knitr::kable(as.matrix(pooled_mq), caption = "Monthly and quarterly cash rate forecasts", digits = 2)
pooled_mq     = merge(pooled_forecasts)
colnames(pooled_mq) = c("monthly", "lower", "upper")
knitr::kable(as.matrix(pooled_mq), caption = "Monthly and quarterly cash rate forecasts", digits = 2)
```

## Survey answers

Based on the forecasts above, and the analysis of forecasts from individual models, I formed the following survey answers:

**When you think the RBA will change the cash rate?**

|          | Jun 2023 | Jul 2023 | Aug 2023 | Sep 2023 | Oct 2023 | Nov 2023 | Dec 2023 | Feb 2024 | March 2024 or later |
|-------|-------|-------|-------|-------|-------|-------|-------|-------|-------|
| Increase | ✓        | ✓        |          |          |          |          |          |          |                   |
| Decrease |          |          |          |          |          |          |          |          |                   |

**Why do you think this?**

> Based on the new data from May, all my forecast systems are aligned and indicate a 15 basis point rise in the cash rate, with further increases expected throughout the year. The forecast intervals have also been narrowed to 3.9 to 4.1%, leaving little doubt about the projected raises. All this seems to be in line with the slowly falling inflation rate, which reached 6.8% in April. My forecasting system for this month includes thirty-two bond yield models for weekly and monthly data and sixty-four complex dynamics time-varying risk models encompassing the leverage effect and risk premium. The forecasts from individual models are further pooled based on their cash rate forecasting capacity.

**By how much do you think the RBA will change the cash rate in the next meeting?**

> 15 pbs

**At what level do you think the cash rate will peak?**

> 4.1%

**When do you think the cash rate will peak?**

> July 2023


## RBA's decision

On 2 May 2023, the RBA announced an **increase** in the cash rate target by **25** basis points.

## Forecasting system

My monthly forecasts were based a similar forecasting system as the one I developed for [April forecasts](2023_04.qmd) as well as a new set of univariate models. The latter includes models of the cash rate capturing complex patterns of data persistence using autoregressive moving average equation extended by time-varying volatility equation - a GARCH model. Part of the models include the leverage effect and/or time-varying risk premium. The best fitting models are with the richest dynamics, that is a (3,3) ARMA order, the GARCH equation with the leverage and in-mean effects. The data is plotted below.

```{r data plot yields}
#| echo: false

plot(x = index(vms), y = as.vector(vms[,2]), main = "Australian interest rates at various maturities",
     type = "l", ylab = "yield [%]", xlab = "time",
     ylim = range(vwa), bty = "n",
     lwd = 1, col = "mediumorchid1"
)
for (i in 3:ncol(vms)) lines(x = index(vms), y = as.vector(vms[,i]), col = cols[i])
lines(x = index(vms), y = as.vector(vms[,1]), col = "darkorchid4", lwd = 2)
legend("topright", legend = colnames(vms), col = cols, lwd = c(2, rep(1, 7)), bty = "n")

```
