---
title: "April 2023"
author: 
  - name: "Tomasz Woźniak"
    affiliations:
      - University of Melbourne
    orcid: 0000-0003-2212-2378
execute:
  freeze: auto
---

```{r interest data}
#| echo: false
#| message: false
#| warning: false

# download daily interest rates
icr_dwnld   = readrba::read_rba(series_id = "FIRMMCRTD")   # Cash Rate Target
icr_tmp     = xts::xts(icr_dwnld$value, icr_dwnld$date)
dates_tmp   = xts::xts(as.Date(icr_dwnld$date), icr_dwnld$date)

by1m_dwnld = readrba::read_rba(series_id = "FIRMMBAB30D")
by1m_tmp   = xts::xts(by1m_dwnld$value, by1m_dwnld$date)

by3m_dwnld = readrba::read_rba(series_id = "FIRMMBAB90D")
by3m_tmp   = xts::xts(by3m_dwnld$value, by3m_dwnld$date)

by6m_dwnld = readrba::read_rba(series_id = "FIRMMBAB180D")
by6m_tmp   = xts::xts(by6m_dwnld$value, by6m_dwnld$date)

by2y_dwnld = readrba::read_rba(series_id = "FCMYGBAG2D")
by2y_tmp   = xts::xts(by2y_dwnld$value, by2y_dwnld$date)

by3y_dwnld = readrba::read_rba(series_id = "FCMYGBAG3D")
by3y_tmp   = xts::xts(by3y_dwnld$value, by3y_dwnld$date)

by5y_dwnld = readrba::read_rba(series_id = "FCMYGBAG5D")
by5y_tmp   = xts::xts(by5y_dwnld$value, by5y_dwnld$date)

by10y_dwnld = readrba::read_rba(series_id = "FCMYGBAG10D")
by10y_tmp   = xts::xts(by10y_dwnld$value, by10y_dwnld$date)

# daily systems

forecast_day              = "/2023-04-26"

variables_all             = na.omit(merge(icr_tmp, by1m_tmp, by3m_tmp, by6m_tmp, by2y_tmp, by3y_tmp, by5y_tmp, by10y_tmp))
colnames(variables_all)   = c("cash rate", "1m", "3m", "6m", "2y", "3y", "5y", "10y")
variables_all             = variables_all[forecast_day]

variables_long            = na.omit(merge(icr_tmp, by2y_tmp, by3y_tmp, by5y_tmp, by10y_tmp))
colnames(variables_long)  = c("cash rate", "2y", "3y", "5y", "10y")
variables_long            = variables_long[forecast_day]

variables_short           = na.omit(merge(icr_tmp, by1m_tmp, by3m_tmp, by6m_tmp))
colnames(variables_short) = c("cash rate", "1m", "3m", "6m")
variables_short           = variables_short[forecast_day]


# weekly and monthly systems
vwa     = xts::to.weekly(variables_all, OHLC = FALSE)
vwl     = xts::to.weekly(variables_long, OHLC = FALSE)
vws     = xts::to.weekly(variables_short, OHLC = FALSE)

vma     = xts::to.monthly(variables_all, OHLC = FALSE)
vml     = xts::to.monthly(variables_long, OHLC = FALSE)
vms     = xts::to.monthly(variables_short, OHLC = FALSE)

# create a dummy for the interest raise regime
T       = nrow(vms)
dm      = xts::xts(as.matrix(rep(0, T)), zoo::index(vms))
dm[(T - 12):T] = 1
colnames(dm) = "dum"
dmf = as.matrix(rep(1,12))
colnames(dmf) = "dum"

Tw      = nrow(vws)
dw      = xts::xts(as.matrix(rep(0, Tw)), zoo::index(vws))
dw[(Tw - 52):Tw] = 1
colnames(dw) = "dum"
dwf = as.matrix(rep(1,55))
colnames(dwf) = "dum"
```

```{r cointegrating rank}
#| echo: false
#| eval: false
#| message: false
#| warning: false
#| results: hide

library(vars)
# Johansen's cointegrating rank test
vecm_vma   = ca.jo(vma, type = "trace", ecdet = "const", K = 5, spec = "transitory")
summary(vecm_vma) # r = 5, N = 8

vecm_vms   = ca.jo(vms, type = "trace", ecdet = "const", K = 5, spec = "transitory")
summary(vecm_vms) # r = 3, N = 4

vecm_vml   = ca.jo(vml, type = "trace", ecdet = "const", K = 5, spec = "transitory")
summary(vecm_vml) # r = 4, N = 5


vecm_vwa   = ca.jo(vwa, type = "trace", ecdet = "const", K = 5, spec = "transitory")
summary(vecm_vwa) # r = 7, N = 8

vecm_vws   = ca.jo(vws, type = "trace", ecdet = "const", K = 5, spec = "transitory")
summary(vecm_vws) # r = 3, N = 4

vecm_vwl   = ca.jo(vwl, type = "trace", ecdet = "const", K = 5, spec = "transitory")
summary(vecm_vwl) # r = 4, N = 5
```

```{r forecasting}
#| echo: false
#| message: false
#| warning: false

library(vars)
# forecast with monthly data
f         = 1
# forecasts = array(NA, c(12, 3, 18))
# loglik    = rep(NA, 18)
forecasts = array(NA, c(12, 3, 16))
loglik    = rep(NA, 16)
vm        = list(vma, vms, vml)
rr        = c(4, 3, 3)

# for (v in 1:3) {
for (v in 2) {
  for (p in c(3, 5, 7, 9, 11, 13, 15, 17)) {
    vecm            = ca.jo(vm[[v]], type = "trace", ecdet = "const", K = p, spec = "transitory", dumvar = dm)
    var_cr          = vec2var(vecm, r = rr[v])
    var_pr          = predict(var_cr, n.ahead = 12, ci = .68, dumvar = dmf)
    forecasts[,,f]  = var_pr$fcst$cash.rate[,1:3]
    loglik[f]       = sum(dnorm(var_cr$resid[,1], log = TRUE))
    f = f + 1
    
    vecm            = ca.jo(vm[[v]], type = "trace", ecdet = "const", K = p, spec = "transitory")
    var_cr          = vec2var(vecm, r = rr[v])
    var_pr          = predict(var_cr, n.ahead = 12, ci = .68)
    forecasts[,,f]  = var_pr$fcst$cash.rate[,1:3]
    loglik[f]       = sum(dnorm(var_cr$resid[,1], log = TRUE))
    f = f + 1
  }
}

ym1 = zoo::as.yearmon("2023-05") # the first forecasted period
ym2 = zoo::as.yearmon("2024-04") # the last forecasted period
s   = seq(ym1, ym2, 1/12) # create yearmon sequence

# weights are proportional to marginal likelihood for the cash rate
ll        = exp(loglik - max(loglik))
weights   = ll/sum(ll)

forecasts_w = forecasts
# for (i in 1:18) {
for (i in 1:16) {
  forecasts_w[,,i]      = weights[i] * forecasts[,,i]
}

# pooled_forecasts_m = apply(forecasts, 1:2, mean)
# colnames(pooled_forecasts_m) = c("forecast", "lower", "upper")
# pooled_forecasts_m = xts::xts(pooled_forecasts_m, s)

pooled_forecasts_m = apply(forecasts_w, 1:2, sum)
colnames(pooled_forecasts_m) = c("forecast", "lower", "upper")
pooled_forecasts_m = xts::xts(pooled_forecasts_m, s)

ym13 = zoo::as.yearmon("2023-04") # forecast origin
s3   = seq(ym13, ym2, 1/12) # create yearmon sequence

ym12 = zoo::as.yearmon("2010-1") # first data point for the plot
s2   = seq(ym12, ym2, 1/12) # create yearmon sequence

datainforecast  = as.vector(vm[[2]][(dim(vm[[2]])[1] - (length(s2) - 12 - 1)):dim(vm[[2]])[1], 1])
last_point      = datainforecast[length(datainforecast)]

cols = c("darkorchid4","mediumorchid1","mediumorchid2","mediumorchid3","hotpink1","hotpink2","hotpink3","hotpink4")



# forecast with weekly data
f           = 1
# forecastsw  = array(NA, c(55, 3, 18))
# loglikw     = rep(NA, 18)
forecastsw  = array(NA, c(55, 3, 16))
loglikw     = rep(NA, 16)
vw          = list(vwa, vws, vwl)
rrw         = c(7, 3, 4)

# for (v in 1:3) {
for (v in 2) {
  for (p in c(3, 5, 7, 9, 11, 13, 15, 17)) {
    vecm            = ca.jo(vw[[v]], type = "trace", ecdet = "const", K = p, spec = "transitory", dumvar = dw)
    var_cr          = vec2var(vecm, r = rrw[v])
    var_pr          = predict(var_cr, n.ahead = 55, ci = .68, dumvar = dwf)
    forecastsw[,,f] = var_pr$fcst$cash.rate[,1:3]
    loglikw[f]      = sum(dnorm(var_cr$resid[,1], log = TRUE))
    f = f + 1
    
    vecm            = ca.jo(vw[[v]], type = "trace", ecdet = "const", K = p, spec = "transitory")
    var_cr          = vec2var(vecm, r = rrw[v])
    var_pr          = predict(var_cr, n.ahead = 55, ci = .68)
    forecastsw[,,f] = var_pr$fcst$cash.rate[,1:3]
    loglikw[f]      = sum(dnorm(var_cr$resid[,1], log = TRUE))
    f = f + 1
  }
}

sw    = as.Date(rep(NA, 55))
sw[1] = as.Date("2023-05-01")
for (i in 2:55) {
  sw[i] = sw[i - 1] + 7
}

# weights are proportional to marginal likelihood for the cash rate
llw        = exp(loglik - max(loglik))
weightsw   = llw/sum(llw)

forecastsw_w = forecastsw
# for (i in 1:18) {
for (i in 1:16) {
  forecastsw_w[,,i]      = weightsw[i] * forecastsw[,,i]
}

pooled_forecasts_ww = apply(forecastsw_w, 1:2, sum)
colnames(pooled_forecasts_ww) = c("forecast", "lower", "upper")
pooled_forecasts_ww = xts::xts(pooled_forecasts_ww, sw)


# pooled_forecasts_w = apply(forecastsw, 1:2, mean)
# colnames(pooled_forecasts_w) = c("forecast", "lower", "upper")
# pooled_forecasts_w = xts::xts(pooled_forecasts_w, sw)
# pooled_forecasts_wm = xts::to.monthly(pooled_forecasts_w, OHLC = FALSE)

pooled_forecasts_wm = xts::to.monthly(pooled_forecasts_ww, OHLC = FALSE)

# pool forecasts
pooled_forecasts = 0.5 * (pooled_forecasts_m + pooled_forecasts_wm)
zoo::write.zoo(pooled_forecasts, sep = ",", file = "forecasts/2023-04.csv")

```

```{r rates data}
#| echo: false
#| message: false
#| warning: false

# download the rates
icr_dwnld   = readrba::read_rba(series_id = "FIRMMCRTD")   # Cash Rate Target
icr_tmp     = xts::xts(icr_dwnld$value, icr_dwnld$date)
dates_tmp   = xts::xts(as.Date(icr_dwnld$date), icr_dwnld$date)
icr         = xts::to.quarterly(icr_tmp, OHLC = FALSE)
icr         = icr[-nrow(icr),]

by10y_dwnld = readrba::read_rba(series_id = "FCMYGBAG10D") # 10-year bonds
by10y_tmp   = xts::xts(by10y_dwnld$value, by10y_dwnld$date)
by10y       = xts::to.quarterly(by10y_tmp, OHLC = FALSE)
by10y       = by10y[-nrow(by10y),]

# "GMAREXPY"  # Survey measure of market economists' inflation expectations; Median inflation for 1 year ahead; Year-ended
piexpe1y_dwnld  = readrba::read_rba(series_id = "GMAREXPY")
piexpe1y_tmp    = xts::xts(piexpe1y_dwnld$value, piexpe1y_dwnld$date, tclass = 'yearqtr')
piexpe1y        = xts::to.quarterly(piexpe1y_tmp, OHLC = FALSE)
  
# "GUNIEXPY"  # Survey measure of union officials' inflation expectations; Median inflation for 1 year ahead; Year-ended
piexpu1y_dwnld  = readrba::read_rba(series_id = "GUNIEXPY")
piexpu1y_tmp    = xts::xts(piexpu1y_dwnld$value, piexpu1y_dwnld$date, tclass = 'yearqtr')
piexpu1y        = xts::to.quarterly(piexpu1y_tmp, OHLC = FALSE)

# "GCPIAGSAQP"  # Consumer price index; All groups; Quarterly change (in per cent)
picpi_dwnld     = readrba::read_rba(series_id = "GCPIAGSAQP")
picpi_tmp       = 4 * xts::xts(c(picpi_dwnld$value, 7.1/4), c(picpi_dwnld$date,"2023-03-31"), tclass = 'yearqtr')
picpi           = xts::to.quarterly(picpi_tmp, OHLC = FALSE)

# "GLFSURSA"    # Unemployed persons as percentage of labour force
ur_dwnld    = readrba::read_rba(series_id = "GLFSURSA")
ur_tmp      = xts::xts(ur_dwnld$value, ur_dwnld$date)
ur          = xts::to.quarterly(ur_tmp, OHLC = FALSE)

# "GLFSEPTPOP"  # Employed persons as percentage of working age civilian population
er_dwnld    = readrba::read_rba(series_id = "GLFSEPTPOP")
er_tmp      = xts::xts(er_dwnld$value, er_dwnld$date)
er          = xts::to.quarterly(er_tmp, OHLC = FALSE)

with_exp            = na.omit(merge(icr, by10y, piexpe1y, piexpu1y, picpi, ur, er))
colnames(with_exp)  = c("cash rate", "10yb", "Epi_eco", "Epi_uni", "pi", "ur", "er")
with_exp            = with_exp["2002-01/2023-03"]

no_exp              = na.omit(merge(icr, by10y, picpi, ur, er))
colnames(no_exp)    = c("cash rate", "10yb", "pi", "ur", "er")
no_exp              = no_exp["/2023-03"]

```

```{r forecasting quarterly}
#| echo: false
#| eval: false
#| message: false
#| warning: false
#| results: hide

# Johansen's test
library(vars)

vecm_we    = ca.jo(with_exp, type = "trace", ecdet = "none", K = 3, spec = "transitory")
summary(vecm_we) # r = 2, N = 7

vecm_noe    = ca.jo(no_exp, type = "trace", ecdet = "none", K = 3, spec = "transitory")
summary(vecm_noe) # r = 2, N = 5


# forecast with quarterly data
f         = 1
forecasts = array(NA, c(4, 3, 2))
vmq       = list(with_exp, no_exp)

for (v in 1:2) {
  vecm            = ca.jo(vmq[[v]], type = "trace", ecdet = "none", K = 3, spec = "transitory")
  var_cr          = vec2var(vecm, r = 2)
  var_pr          = predict(var_cr, n.ahead = 4, ci = .68)
  forecasts[,,f]  = var_pr$fcst$cash.rate[,1:3]
  f = f + 1
}

ym1q = zoo::as.yearqtr("2023-02") # the first forecasted period
ym2q = zoo::as.yearqtr("2024-01") # the last forecasted period
sq   = seq(ym1, ym2, 1/4) # create yearqtr sequence

pooled_forecasts_q = apply(forecasts, 1:2, mean)
colnames(pooled_forecasts_q) = c("forecast", "lower", "upper")
pooled_forecasts_q = xts::xts(pooled_forecasts_q, sq + 1/12)

ym13q = zoo::as.yearqtr("2023-01") # forecast origin
s3q   = seq(ym13q, ym2q, 1/4) # create yearmon sequence

ym12q = zoo::as.yearqtr("2010-01")
s2q   = seq(ym12q, ym2q, 1/4) # create yearmon sequence

datainforecastq  = as.vector(vmq[[1]][(dim(vmq[[1]])[1] - (length(s2q) - 4 - 1)):dim(vmq[[1]])[1], 1])
last_pointq      = datainforecastq[length(datainforecastq)]

cols = c("darkorchid4","mediumorchid1","mediumorchid2","mediumorchid3","hotpink1","hotpink2","hotpink3","hotpink4")

zoo::write.zoo(pooled_forecasts_q, sep = ",", file = "forecasts/2023-04-quarterly.csv")

```

> The end-of-April forecasting for the RBA cash rate survey by [finder.com.au](https://www.finder.com.au/rba-cash-rate) follows the anouncement of inflation for the first quarter of 2023 at the level of 7%. It fell, but remains high. The new data leads to forecasts indicating a slight increase in the cash rate for May. In the longer horizon, the bond yield curve modelling indicates stabilisation or further increases of cash rate while a system focusing on inflation and its expectations show its decline.

## Cash rate forecasts

The figure below presents the monthly cash rate series starting from January 2010, with the forecasts reported from May 2023 to April 2024 as the forecast mean and the 68% forecasting intervals.

```{r forecast plot}
#| echo: false


ci1_tmp = col2rgb(cols[2])
ci2_tmp = col2rgb(cols[4])
ci1     = rgb(ci1_tmp[1], ci1_tmp[2], ci1_tmp[3], 100, maxColorValue = 255)
ci2     = rgb(ci2_tmp[1], ci2_tmp[2], ci2_tmp[3], 100, maxColorValue = 255)

plot(x = s2, y = c(datainforecast, pooled_forecasts[,1]), main = "Cash rate forecast",
     type = "l", ylab = "[%]", xlab = "time",
     ylim = range(pooled_forecasts, datainforecast), bty = "n",
     lwd = 1, col = cols[1]
)
polygon(x = c(s3, s3[13:1]), 
        y = c(last_point,as.vector(pooled_forecasts[,2]), as.vector(pooled_forecasts[,3])[12:1], last_point),
        col = ci1, border = ci1)
# polygon(x = c(s3q + .25, s3q[5:1] + .25), 
#         y = c(last_pointq,as.vector(pooled_forecasts_q[,2]), as.vector(pooled_forecasts_q[,3])[4:1], last_pointq),
#         col = ci2, border = ci2)
lines(x = s2, y = c(datainforecast, pooled_forecasts[,1]), lwd = 2, col = cols[1])
# lines(x = s3q + .25, y = c(last_pointq, pooled_forecasts_q[,1]), lwd = 2, col = cols[4])
abline(v = ym13, col = cols[6], lty = 3)


```

The table below makes the numerical values presented in the figure more accessible.

```{r forecast table}
#| echo: false
options(knitr.kable.NA = '') 
# pooled_mq     = merge(pooled_forecasts,pooled_forecasts_q)
# colnames(pooled_mq) = c("monthly", "lower", "upper", "quarterly", "lower", "upper")
# knitr::kable(as.matrix(pooled_mq), caption = "Monthly and quarterly cash rate forecasts", digits = 2)
pooled_mq     = merge(pooled_forecasts)
colnames(pooled_mq) = c("monthly", "lower", "upper")
knitr::kable(as.matrix(pooled_mq), caption = "Monthly and quarterly cash rate forecasts", digits = 2)
```

## Survey answers

Based on the forecasts above, and the analysis of forecasts from individual models, I formed the following survey answers:

**When you think the RBA will change the cash rate?**

|          | May 2023 | Jun 2023 | Jul 2023 | Aug 2023 | Sep 2023 | Oct 2023 | Nov 2023 | Dec 2023 | Jan 2024 or later |
|-------|-------|-------|-------|-------|-------|-------|-------|-------|-------|
| Increase | ✓        | ✓        |          |          |          |          |          |          |                   |
| Decrease |          |          |          |          |          |          |          |          |                   |

**Why do you think this?**

> The new reading of the CPI inflation reaching 7% in the first quarter of 2023 aligns the quarterly forecasts with the monthly indicating the cash rate at 3.77% within the next 3-5 months. The prediction bands around this value reach from 3.4 to 4.1 %. My interpretation is a likely raise by 15 pp this or next month. Beyond this horizon, the forecasts diverge with the bond yield curve modelling indicating further increases and the system focusing on inflation, labour market and expectations showing cuts. The arrival of new macro data occurs essential in such dynamically changing circumstances./

**By how much do you think the RBA will change the cash rate in the next meeting?**

> 15 pbs

**At what level do you think the cash rate will peak?**

> 3.8%

**When do you think the cash rate will peak?**

> June 2023

**Where do you think the cash rate will be at the end of 2023?**

> 3.8%

## RBA's decision

On 2 May 2023, the RBA announced an **increase** in the cash rate target by **25** basis points.

## Forecasting system

<!-- My quarterly forecasts are based on the same system that presented in [January forecasts](2023_01.qmd). The quarterly data spanning the period starting in the second quarter of 2002 and finishing in the first one of 2023 is plotted below. -->

```{r data plot rates}
#| echo: false
#| eval: false

plot(x = index(with_exp), y = as.vector(with_exp[,1]), main = "Quarterly data",
     type = "l", ylab = "[%]", xlab = "time",
     ylim = range(with_exp[,1:6], with_exp[,7]-60), bty = "n",
     lwd = 2, col = "darkorchid4"
)
for (i in 2:6) lines(x = index(with_exp), y = as.vector(with_exp[,i]), col = cols[i])
lines(x = index(with_exp), y = as.vector(with_exp[,7]) - 60, col = cols[7], lwd = 2)
lines(x = index(with_exp), y = as.vector(with_exp[,1]), col = "darkorchid4", lwd = 2)
axis(2, lwd = 2, col = "darkorchid4")
axis(4, c(-1, 2, 5), c(59, NA, 65), lwd = 2, col = cols[7])

var_names = c("cash rate", "10y bond yield", "inflation expectation eco", "inflation expectation union", "inflation", "unemployment", "employment")
legend("bottomleft", legend = var_names, col = cols[1:7], lwd = c(2, rep(1, 5), 2), bty = "n")
```

My monthly forecasts were based a similar forecasting system as the one I developed for [March forecasts](2023_03.qmd). However, due to [unavailability of long-term interest rates via package](https://github.com/MattCowgill/readrba/issues/43) **readrba** I only used cash rate and government bond yields at maturity 30, 90, and 180 days. The data is plotted below.

```{r data plot yields}
#| echo: false

plot(x = index(vms), y = as.vector(vms[,2]), main = "Australian interest rates at various maturities",
     type = "l", ylab = "yield [%]", xlab = "time",
     ylim = range(vwa), bty = "n",
     lwd = 1, col = "mediumorchid1"
)
for (i in 3:ncol(vms)) lines(x = index(vms), y = as.vector(vms[,i]), col = cols[i])
lines(x = index(vms), y = as.vector(vms[,1]), col = "darkorchid4", lwd = 2)
legend("topright", legend = colnames(vms), col = cols, lwd = c(2, rep(1, 7)), bty = "n")

```
