---
title: "June 2023"
author: 
  - name: "Tomasz Woźniak"
    url: https://github.com/donotdespair
    affiliations:
      - University of Melbourne
    orcid: 0000-0003-2212-2378
execute:
  freeze: auto
---

```{r interest data}
#| echo: false
#| message: false
#| warning: false

# download daily interest rates
icr_dwnld   = readrba::read_rba(series_id = "FIRMMCRTD")   # Cash Rate Target
icr_tmp     = xts::xts(icr_dwnld$value, icr_dwnld$date)
dates_tmp   = xts::xts(as.Date(icr_dwnld$date), icr_dwnld$date)

by1m_dwnld = readrba::read_rba(series_id = "FIRMMBAB30D")
by1m_tmp   = xts::xts(by1m_dwnld$value, by1m_dwnld$date)

by3m_dwnld = readrba::read_rba(series_id = "FIRMMBAB90D")
by3m_tmp   = xts::xts(by3m_dwnld$value, by3m_dwnld$date)

by6m_dwnld = readrba::read_rba(series_id = "FIRMMBAB180D")
by6m_tmp   = xts::xts(by6m_dwnld$value, by6m_dwnld$date)

by2y_dwnld = readrba::read_rba(series_id = "FCMYGBAG2D")
by2y_tmp   = xts::xts(by2y_dwnld$value, by2y_dwnld$date)

by3y_dwnld = readrba::read_rba(series_id = "FCMYGBAG3D")
by3y_tmp   = xts::xts(by3y_dwnld$value, by3y_dwnld$date)

by5y_dwnld = readrba::read_rba(series_id = "FCMYGBAG5D")
by5y_tmp   = xts::xts(by5y_dwnld$value, by5y_dwnld$date)

by10y_dwnld = readrba::read_rba(series_id = "FCMYGBAG10D")
by10y_tmp   = xts::xts(by10y_dwnld$value, by10y_dwnld$date)

long_ou_tmp = na.omit(merge(by2y_tmp, by3y_tmp, by5y_tmp, by10y_tmp))
long_be     = long_ou_tmp["/2013-05-16"]
long_af     = long_ou_tmp["2022-01-01/"]

long_in_tmp = readxl::read_xls(path = "f02d.xls", skip = 10)
long_in     = xts::xts(long_in_tmp[,2:5], as.Date(long_in_tmp$`Series ID`))
long_in     = long_in["2013-05-17/2021-12-31"]
colnames(long_in) <- colnames(long_af)

short       = na.omit(merge(icr_tmp, by1m_tmp, by3m_tmp, by6m_tmp))
long        = rbind(long_be, long_in, long_af)

# daily systems
forecast_day              = "/2023-06-28"

variables_all             = na.omit(merge(short, long))
colnames(variables_all)   = c("cash rate", "1m", "3m", "6m", "2y", "3y", "5y", "10y")
variables_all             = variables_all[forecast_day]

variables_long            = na.omit(merge(icr_tmp, long))
colnames(variables_long)  = c("cash rate", "2y", "3y", "5y", "10y")
variables_long            = variables_long[forecast_day]

variables_short           = short
colnames(variables_short) = c("cash rate", "1m", "3m", "6m")
variables_short           = variables_short[forecast_day]


# weekly and monthly systems
vwa     = xts::to.weekly(variables_all, OHLC = FALSE)
vwl     = xts::to.weekly(variables_long, OHLC = FALSE)
vws     = xts::to.weekly(variables_short, OHLC = FALSE)

vma     = xts::to.monthly(variables_all, OHLC = FALSE)
vml     = xts::to.monthly(variables_long, OHLC = FALSE)
vms     = xts::to.monthly(variables_short, OHLC = FALSE)

# create a dummy for the interest raise regime
T       = nrow(vms)
dm      = xts::xts(as.matrix(rep(0, T)), zoo::index(vms))
dm["2022-05/"] = 1
colnames(dm) = "dum"
dmf = as.matrix(rep(1,12))
colnames(dmf) = "dum"

Tw      = nrow(vws)
dw      = xts::xts(as.matrix(rep(0, Tw)), zoo::index(vws))
dw["2022-05/"] = 1
colnames(dw) = "dum"
dwf = as.matrix(rep(1,55))
colnames(dwf) = "dum"
```

```{r cointegrating rank}
#| echo: false
#| eval: false
#| message: false
#| warning: false
#| results: hide

library(vars)
# Johansen's cointegrating rank test
vecm_vma   = ca.jo(vma, type = "trace", ecdet = "const", K = 5, spec = "transitory")
summary(vecm_vma) # r = 5, N = 8

vecm_vms   = ca.jo(vms, type = "trace", ecdet = "const", K = 5, spec = "transitory")
summary(vecm_vms) # r = 3, N = 4

vecm_vml   = ca.jo(vml, type = "trace", ecdet = "const", K = 5, spec = "transitory")
summary(vecm_vml) # r = 4, N = 5


vecm_vwa   = ca.jo(vwa, type = "trace", ecdet = "const", K = 5, spec = "transitory")
summary(vecm_vwa) # r = 7, N = 8

vecm_vws   = ca.jo(vws, type = "trace", ecdet = "const", K = 5, spec = "transitory")
summary(vecm_vws) # r = 3, N = 4

vecm_vwl   = ca.jo(vwl, type = "trace", ecdet = "const", K = 5, spec = "transitory")
summary(vecm_vwl) # r = 4, N = 5
```

```{r forecasting}
#| echo: false
#| message: false
#| warning: false

library(vars)
# forecast with monthly data
f         = 1
vm        = list(vma, vms, vml)
pp        = c(3, 5, 7, 9, 11, 13, 15, 17)
rr        = c(5, 3, 4)
mm        = length(pp) * length(vm)
forecasts = array(NA, c(12, 3, mm))
loglik    = rep(NA, mm)

for (v in 1:3) {
  for (p in pp) {
    # vecm            = ca.jo(vm[[v]], type = "trace", ecdet = "const", K = p, spec = "transitory", dumvar = dm)
    # var_cr          = vec2var(vecm, r = rr[v])
    # var_pr          = predict(var_cr, n.ahead = 12, ci = .68, dumvar = dmf)
    # forecasts[,,f]  = var_pr$fcst$cash.rate[,1:3]
    # loglik[f]       = sum(dnorm(var_cr$resid[,1], log = TRUE))
    # f = f + 1
    
    vecm            = ca.jo(vm[[v]], type = "trace", ecdet = "const", K = p, spec = "transitory")
    var_cr          = vec2var(vecm, r = rr[v])
    var_pr          = predict(var_cr, n.ahead = 12, ci = .68)
    forecasts[,,f]  = var_pr$fcst$cash.rate[,1:3]
    loglik[f]       = sum(dnorm(var_cr$resid[,1], log = TRUE))
    f = f + 1
  }
}

ym13 = zoo::as.yearmon("2023-06") # forecast origin
ym1 = ym13 + 1/12                 # the first forecasted period
ym2 = ym1 + 11/12                 # the last forecasted period
s   = seq(ym1, ym2, 1/12)         # create yearmon sequence

# weights are proportional to marginal likelihood for the cash rate
ll        = exp(loglik - max(loglik))
weights   = ll/sum(ll)

forecasts_w = forecasts
for (i in 1:mm) {
  forecasts_w[,,i]      = weights[i] * forecasts[,,i]
}

pooled_forecasts_m = apply(forecasts_w, 1:2, sum)
colnames(pooled_forecasts_m) = c("forecast", "lower", "upper")
pooled_forecasts_m = xts::xts(pooled_forecasts_m, s)

s3   = seq(ym13, ym2, 1/12) # create yearmon sequence
ym12 = zoo::as.yearmon("2010-1") # first data point for the plot
s2   = seq(ym12, ym2, 1/12) # create yearmon sequence

datainforecast  = as.vector(vm[[2]][(dim(vm[[2]])[1] - (length(s2) - 12 - 1)):dim(vm[[2]])[1], 1])
last_point      = datainforecast[length(datainforecast)]

cols = c("darkorchid4","mediumorchid1","mediumorchid2","mediumorchid3","hotpink1","hotpink2","hotpink3","hotpink4")



# forecast with weekly data
f           = 1
vw          = list(vwa, vws, vwl)
rrw         = c(7, 3, 4)
mmw         = length(pp) * length(vw)
forecastsw  = array(NA, c(55, 3, mmw))
loglikw     = rep(NA, mmw)

for (v in 1:3) {
  for (p in pp) {
    # vecm            = ca.jo(vw[[v]], type = "trace", ecdet = "const", K = p, spec = "transitory", dumvar = dw)
    # var_cr          = vec2var(vecm, r = rrw[v])
    # var_pr          = predict(var_cr, n.ahead = 55, ci = .68, dumvar = dwf)
    # forecastsw[,,f] = var_pr$fcst$cash.rate[,1:3]
    # loglikw[f]      = sum(dnorm(var_cr$resid[,1], log = TRUE))
    # f = f + 1
    
    vecm            = ca.jo(vw[[v]], type = "trace", ecdet = "const", K = p, spec = "transitory")
    var_cr          = vec2var(vecm, r = rrw[v])
    var_pr          = predict(var_cr, n.ahead = 55, ci = .68)
    forecastsw[,,f] = var_pr$fcst$cash.rate[,1:3]
    loglikw[f]      = sum(dnorm(var_cr$resid[,1], log = TRUE))
    f = f + 1
  }
}

sw    = as.Date(rep(NA, 55))
sw[1] = as.Date("2023-07-05") # the first forecast date
for (i in 2:55) {
  sw[i] = sw[i - 1] + 7
}

# weights are proportional to marginal likelihood for the cash rate
llw        = exp(loglikw - max(loglikw))
weightsw   = llw/sum(llw)

forecastsw_w = forecastsw
for (i in 1:mmw) {
  forecastsw_w[,,i]      = weightsw[i] * forecastsw[,,i]
}

pooled_forecasts_ww = apply(forecastsw_w, 1:2, sum)
colnames(pooled_forecasts_ww) = c("forecast", "lower", "upper")
pooled_forecasts_ww = xts::xts(pooled_forecasts_ww, sw)


# pooled_forecasts_w = apply(forecastsw, 1:2, mean)
# colnames(pooled_forecasts_w) = c("forecast", "lower", "upper")
# pooled_forecasts_w = xts::xts(pooled_forecasts_w, sw)
# pooled_forecasts_wm = xts::to.monthly(pooled_forecasts_w, OHLC = FALSE)

pooled_forecasts_wm = xts::to.monthly(pooled_forecasts_ww, OHLC = FALSE)

```

```{r univariate forecasting}
#| echo: false
#| message: false
#| warning: false

crm = vms[,1]
crw = vws[,1]

# monthly arma-garch forecasting
p_max     = 3
q_max     = 3

garchfm   = array(NA, c(12, 3, 4, (p_max ) * (q_max + 1)))
loglik    = matrix(NA, 4, (p_max ) * (q_max + 1))

for (i in 1:p_max){
  for (j in 0:q_max){
    # arma-garch forecasting
    garch_ms   = rugarch::ugarchspec(
      variance.model = list(model = "sGARCH"), 
      mean.model = list(armaOrder = c(i, j)), 
      distribution.model = "norm"
    )
    garch_mf   = rugarch::ugarchfit(garch_ms, crm, solver = "hybrid")
    garch_mfor = rugarch::ugarchboot(garch_mf, n.ahead = 12,
                                     method = "Partial", n.bootpred = 5000)
    garchfm[,,1, (q_max + 1)*(i - 1) + j + 1]  = cbind(garch_mfor@forc@forecast$seriesFor,
          t(apply(garch_mfor@fseries, 2, HDInterval::hdi , credMass = .68)))
    loglik[1, (q_max + 1)*(i - 1) + j + 1]     = -sum(garch_mf@fit$log.likelihoods)
    
    # arma-garch-in-mean forecasting
    garch_ms   = rugarch::ugarchspec(
      variance.model = list(model = "sGARCH"), 
      mean.model = list(armaOrder = c(i, j), archm = TRUE), 
      distribution.model = "norm"
    )
    garch_mf   = rugarch::ugarchfit(garch_ms, crm, solver = "hybrid")
    garch_mfor = rugarch::ugarchboot(garch_mf, n.ahead = 12,
                                     method = "Partial", n.bootpred = 5000)
    garchfm[,,2, (q_max + 1)*(i - 1) + j + 1]  = cbind(garch_mfor@forc@forecast$seriesFor,
          t(apply(garch_mfor@fseries, 2, HDInterval::hdi , credMass = .68)))
    loglik[2, (q_max + 1)*(i - 1) + j + 1]     = -sum(garch_mf@fit$log.likelihoods)
    
    # arma-gjr-garch forecasting
    garch_ms   = rugarch::ugarchspec(
      variance.model = list(model = "gjrGARCH"), 
      mean.model = list(armaOrder = c(i, j)), 
      distribution.model = "norm"
    )
    garch_mf   = rugarch::ugarchfit(garch_ms, crm, solver = "hybrid")
    garch_mfor = rugarch::ugarchboot(garch_mf, n.ahead = 12,
                                     method = "Partial", n.bootpred = 5000)
    garchfm[,,3, (q_max + 1)*(i - 1) + j + 1]  = cbind(garch_mfor@forc@forecast$seriesFor,
          t(apply(garch_mfor@fseries, 2, HDInterval::hdi , credMass = .68)))
    loglik[3, (q_max + 1)*(i - 1) + j + 1]     = -sum(garch_mf@fit$log.likelihoods)
    
    # arma-gjr-garch-in-mean forecasting
    garch_ms   = rugarch::ugarchspec(
      variance.model = list(model = "gjrGARCH"), 
      mean.model = list(armaOrder = c(i, j), archm = TRUE), 
      distribution.model = "norm"
    )
    garch_mf   = rugarch::ugarchfit(garch_ms, crm, solver = "hybrid")
    garch_mfor = rugarch::ugarchboot(garch_mf, n.ahead = 12,
                                     method = "Partial", n.bootpred = 5000)
    garchfm[,,4, (q_max + 1)*(i - 1) + j + 1]  = cbind(garch_mfor@forc@forecast$seriesFor,
          t(apply(garch_mfor@fseries, 2, HDInterval::hdi , credMass = .68)))
    loglik[4, (q_max + 1)*(i - 1) + j + 1]     = -sum(garch_mf@fit$log.likelihoods)
  }
}
loglik[is.na(loglik)] = -1000000

llm        = exp(loglik - max(loglik))
weightsm   = llm/sum(llm)

forecastsm = garchfm
forecastsm[is.na(forecastsm)] = 0
for (i in 1:dim(garchfm)[3]) {
  for (j in 1:dim(garchfm)[4]){
    forecastsm[,,i,j]      = weightsm[i,j] * forecastsm[,,i,j]
  }
}

pooled_forecasts_garchm = apply(forecastsm, 1:2, sum)
colnames(pooled_forecasts_garchm) = c("forecast", "lower", "upper")
pooled_forecasts_garchm = xts::xts(pooled_forecasts_garchm, s)

```

```{r pool forecasts}
#| echo: false
#| message: false
#| warning: false

# pool forecasts
pooled_forecasts = (1/3) * (pooled_forecasts_m + pooled_forecasts_wm + pooled_forecasts_garchm)
zoo::write.zoo(pooled_forecasts, sep = ",", file = "forecasts/2023-06.csv")

```

> The end-of-May forecasting for the RBA cash rate survey by [finder.com.au](https://www.finder.com.au/rba-cash-rate) follows the anouncement of inflation for June 2023 at the level of 5.6%. The new data leads to surprising forecasts indicating a decisive increase in the cash rate in July and subsequent increases further on.

## Cash rate forecasts

The figure below presents the monthly cash rate series starting from January 2010, with the forecasts reported from July 2023 to June 2024 as the forecast mean and the 68% forecasting intervals.

```{r forecast plot}
#| echo: false


ci1_tmp = col2rgb(cols[2])
ci2_tmp = col2rgb(cols[4])
ci1     = rgb(ci1_tmp[1], ci1_tmp[2], ci1_tmp[3], 100, maxColorValue = 255)
ci2     = rgb(ci2_tmp[1], ci2_tmp[2], ci2_tmp[3], 100, maxColorValue = 255)

plot(x = s2, y = c(datainforecast, pooled_forecasts[,1]), main = "Cash rate forecast",
     type = "l", ylab = "[%]", xlab = "time",
     ylim = range(pooled_forecasts, datainforecast), bty = "n",
     lwd = 1, col = cols[1]
)
polygon(x = c(s3, s3[13:1]), 
        y = c(last_point,as.vector(pooled_forecasts[,2]), as.vector(pooled_forecasts[,3])[12:1], last_point),
        col = ci1, border = ci1)
lines(x = s2, y = c(datainforecast, pooled_forecasts[,1]), lwd = 2, col = cols[1])
abline(v = ym13, col = cols[6], lty = 3)


```

The table below makes the numerical values presented in the figure more accessible.

```{r forecast table}
#| echo: false
options(knitr.kable.NA = '') 
# pooled_mq     = merge(pooled_forecasts,pooled_forecasts_q)
# colnames(pooled_mq) = c("monthly", "lower", "upper", "quarterly", "lower", "upper")
# knitr::kable(as.matrix(pooled_mq), caption = "Monthly and quarterly cash rate forecasts", digits = 2)
pooled_mq     = merge(pooled_forecasts)
colnames(pooled_mq) = c("monthly", "lower", "upper")
knitr::kable(as.matrix(pooled_mq), caption = "Monthly and quarterly cash rate forecasts", digits = 2)
```

## Survey answers

Based on the forecasts above, and the analysis of forecasts from individual models, I formed the following survey answers:

**When you think the RBA will change the cash rate?**

|          | Jul 2023 | Aug 2023 | Sep 2023 | Oct 2023 | Nov 2023 | Dec 2023 | Feb 2024 | Mar 2024 | Apr 2024 or later |
|--------|--------|--------|--------|--------|--------|--------|--------|--------|--------|
| Increase | ✓        | ✓        |          |          |          |          |          |          |                   |
| Decrease |          |          |          |          |          |          |          |          |                   |

**Why do you think this?**

> Despite the lower reading of monthly inflation in June at 5.3%, my forecasts tell the same story as last month. They indicate a 15 basis point rise in the cash rate, with further increases expected throughout the year. The narrow forecast interval, spanning the values from 4.16 to 4.34%, leaves little doubt about the projected raises.

**By how much do you think the RBA will change the cash rate in the next meeting?**

> 15 pbs

**At what level do you think the cash rate will peak?**

> 4.25%

**When do you think the cash rate will peak?**

> August 2023

## RBA's decision

On 4 July 2023, the RBA announced **no change** in the cash rate target.

## Forecasting system

My forecasting system for June is based on the cash rate target and government bond yields at various maturities as presented in the figure below.

```{r data plot yields}
#| echo: false

plot(x = index(vwa), y = as.vector(vwa[,2]), main = "Australian interest rates at various maturities",
     type = "l", ylab = "yield [%]", xlab = "time",
     ylim = range(vwa), bty = "n",
     lwd = 1, col = "mediumorchid1"
)
for (i in 3:8) lines(x = index(vwa), y = as.vector(vwa[,i]), col = cols[i])
lines(x = index(vwa), y = as.vector(vwa[,1]), col = "darkorchid4", lwd = 2)
legend("topright", legend = colnames(vwa), col = cols, lwd = c(2, rep(1, 7)), bty = "n")

```

The system consists of a hundred of models. Half of them are Vector Error Correction models for weekly and monthly series with different model specification parameters. This month, for the first time, I did not use models with the dummy variable for th recent cash rate hike period and assuming this regime over all forecasting horizons.

The other half of models includes univariate specifications for the cash rate capturing complex patterns of data persistence using autoregressive moving average equation extended by time-varying volatility equation - a GARCH model. Part of the models include the leverage effect and/or time-varying risk premium. The best fitting models are with the richest dynamics, that is a (3,3) ARMA order, the GARCH equation with the leverage and in-mean effects. The data is plotted below.
